---
title: "Assignment 1: Food Delivery Time Analysis"
author: "Daniel Drasler"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Data Source

**Food Delivery Time Prediction Data set** — 50,000 simulated food delivery
orders with 24 variables.

## 1. Import the Dataset

```{r import}
delivery <- read.csv("Food_Delivery_Time_Prediction.csv")
str(delivery)
```

The dataset contains 50,000 rows and 24 columns, with no missing values in
any column. It includes both quantitative variables (delivery time, road
distance, preparation time) and categorical variables (weather, traffic
level, vehicle type).

## 2. Descriptive Statistics — Quantitative Variables

Three quantitative variables were selected: delivery time, road distance,
and preparation time.

```{r desc-quant}
summary(delivery[, c("Time_taken_min", "Road_Distance_km", "Preparation_Time_Min")])
```

Delivery time averages roughly 84 minutes with a median around 78. The mean
sitting above the median suggests the distribution is right-skewed, which is
confirmed in the histogram below.

## 3. Descriptive Statistics — Categorical Variables

Two categorical variables were selected: traffic level and weather.

```{r desc-cat}
table(delivery$Traffic_Level)
table(delivery$Weather)

# As percentages
round(100 * prop.table(table(delivery$Traffic_Level)), 1)
```

Most deliveries occur under low-to-moderate traffic and clear weather, with
"Severe" traffic and "Storm" conditions being comparatively rare categories.

A useful cross-check is comparing the quantitative variable across these
categorical groups:

```{r desc-cross}
round(tapply(delivery$Time_taken_min, delivery$Traffic_Level, mean), 1)
```

Average delivery time rises steadily with traffic level, from roughly 76
minutes under low traffic to over 130 minutes under severe traffic.

## 4. Variable Transformations

**Transformation 1 — log transform (quantitative).** Delivery time is
right-skewed, so a natural-log transformation was applied to compress the
long upper tail.

```{r transform-log}
delivery$log_time <- log(delivery$Time_taken_min)
head(delivery[, c("Time_taken_min", "log_time")])
```

**Transformation 2 — level collapsing (categorical).** The five-level
`Weather` variable was collapsed into a simpler two-level indicator of
adverse versus clear conditions.

```{r transform-cat}
delivery$adverse_weather <- factor(
  ifelse(delivery$Weather %in% c("Rain", "Storm", "Fog"), "Adverse", "Clear/Cloudy")
)
table(delivery$adverse_weather)
```

About 30% of deliveries (14,923 of 50,000) occurred in adverse weather.

## 5. Plots

**Histogram of a quantitative variable.**

```{r plot-hist, fig.width=7, fig.height=5}
hist(delivery$Time_taken_min,
     main = "Distribution of Delivery Time",
     xlab = "Delivery time (minutes)",
     col  = "steelblue")
```

The histogram confirms the right skew noted earlier: most deliveries cluster
between roughly 60 and 90 minutes, with a tail extending out toward the
180-minute ceiling.

**Scatterplot.**

```{r plot-scatter, fig.width=7, fig.height=5}
plot(delivery$Road_Distance_km, delivery$Time_taken_min,
     main = "Delivery Time vs. Road Distance",
     xlab = "Road distance (km)",
     ylab = "Delivery time (minutes)",
     pch  = 19,
     col  = rgb(0.2, 0.4, 0.7, 0.2))
```

There is a clear positive relationship between road distance and delivery
time.. Farther deliveries take longer, as expected. The flat band of points
along the top of the plot (delivery time capped near 180 minutes) is visible
here as well. Overall, it reflects a ceiling in how the data was generated rather than a naturally occurring pattern in delivery logistics.

## Summary

This analysis imported a 50,000-row food delivery data set, summarized three
quantitative and two categorical variables, applied a numeric (log) and a
categorical (level-collapsing) transformation, and produced a histogram and
a scatter plot. The clearest finding is that both traffic level and distance
are strongly associated with delivery time, while the 180-minute cap in the
data is a generation artifact worth noting rather than a real-world limit.
